Measuring contrast processing in the visual system using SSVEP

### 2024-05-05 review.tex
Added more figure cation, grids
Editing attention

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, eccentricity, chromaticity
and the presence of other stimuli has been a focus of vision science for
at least 100 years. One of the most productive experimental approaches
in this field has been the use of the `steady-state visually-evoked
potential' (SSVEP): a technique where contrast modulating inputs are
`frequency tagged' (presented at well-defined frequencies and phases)
and the electrical signals that they generate in the brain are analyzed
in the temporal frequency domain. SSVEPs have several advantages over
conventional measures of visually-evoked responses: they have relatively
unambiguous ouput measures, a high SNR, and they allow us to analyze
interactions between stimulus components using a convenient mathematical
framework. Here we describe how SSVEPs have been used to study visual
contrast over the past 70 years (Dawson, 1954). Because our thinking
about SSVEPs is well-described by simple mathematical models, we embed
code that illustrates key steps in the modelling and analysis. This
paper can therefore be used both as a review of the use of SSVEP in
measuring human contrast processing, and as an interactive learning aid.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic contrast
and although most of the research we describe here focuses on achromatic
contrast, SSVEPs have proven to be an excellent measure of early
chromatic processing as well (McKeefry \emph{et al.}, 1996; Baseler \&
Sutter, 1997; Di Russo \emph{et al.}, 2001).

Contrast is relatively simple to define: typically, contrast is
specified as the percentage deviation of a uniform stimulus from the
background. So, for example, a disk of 100 units of cone activation
(\(I_{\mathrm{stim}}\)) surrounded by a `background' of 50 units of
activation (\(I_{\mathrm{background}}\)) has a contrast of
\(\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}\)
= 100\%. Where patterns are more complex (for example, the sine-wave
gratings or Gabor patches common in vision science), the Michelson
(1927) definition of contrast is specified by the maximum and minimum
excursions from the mean:

\begin{equation}
\frac {I_{\mathrm{stimmax} }-I_{\mathrm{stimmin} }}{I_{\mathrm{stimmax} }+I_{\mathrm{stimmin} }}.
\end{equation}

@@

\includegraphics{review_files/figure-pdf/fig-centresurround-output-1.pdf}

@@

\caption{\label{fig-centresurround}The perceived contrast of a stimulus
depends on its context. A high contrast surround reduces the apparent
contrast of the central `probe' region''}

@@

Because it provides a direct read-out of neural population activity, the
SSVEP signal can reveal key features of neural signal transduction. For
example, by varying the peak stimulus contrast parametrically, a
`contrast vs response' function can be measured - where the `response'
is typically defined as the amplitude of the SSVEP frequency component
at the stimulus frequency or a low multiple thereof. This corresponds
closely to similar functions reported by studies measuring single unit
activity or local field potentials in the cortex (Shapley \& Victor,
1980; Morrone \emph{et al.}, 1982). However the SSVEP has the advantage
that it is non-invasive, and so can be measured in awake, behaving human
participants.

@@

Previous figures showed how a single-input SSVEP might generate a single
set of peaks in the frequency domain response which, in turn, can be
used as a read-out of neuronal activity. In
Figure~\ref{fig-plotspectra}, we show the effect of adding a second
contrast component: the two components interact generating nonlinear
intermodulation terms at sums and differences of the input frequencies
(F1, F2). The nature of the interaction --- and therefore the pattern of
intermodulation terms --- is determined by the computations happening as
the contrast signal moves from retina to cortex (Regan \& Regan, 1988)

First, we implement a simple non-linear transducer with a single input
to illustrate the typical sigmoidal shape of a contrast response
function (Figure~\ref{fig-contrastresponsefunction}).

Next, we show that adding a second component which contributes to the
gain pool of the first, results in a complex pattern of intermodulation
terms (Figure~\ref{fig-plotspectra}).

@@

\includegraphics{review_files/figure-pdf/fig-plotsinewaves-output-1.pdf}

@@

\caption{\label{fig-plotsinewaves}a) A typical SSVEP setup: a screen, a
grating, a flicker rate: b) The responses from a single electrode (time
domain) c) Frequency domain}

@@

The two inputs are represented here as amplitude modulations across
time.

If the two inputs are simply added together, the representation of the
resulting signal in the Fourier domain is just the linear sum of the two
independent signals (first two panels of Figure~\ref{fig-plotspectra}).
However, it is important to take physiology into account. If we ignore
cells sensitive to contrast transients, contrast is represented as a
combination of rectified signals from `on' and `off'-sensitive cells
(broadly, cells that code positive or negative changes in spatiotemporal
luminance). To model the population response of these cells to a
contrast-reversing luminance sine wave, we therefore introduce an
initial full-wave rectification stage that simulates the outputs of two,
equal populations of on- and off-polarity detectors. Once this is done,
we see that the Fourier spectrum of the resulting signal contains power
only at the second harmonics and above (final two panels of
Figure~\ref{fig-plotspectra}). It is possible to generate first-hamonic
responses in contrast-response SSVEPs, but only when the inputs are
modulated in an `on/off' manner rather than contrast reversed.

@@

\includegraphics{review_files/figure-pdf/fig-plotspectra-output-1.pdf}

@@

\caption{\label{fig-plotspectra}Temporal frequency representations of
combining two inputs at frequencies F1=5Hz and F2=7Hz combination. A
pure sine-wave input would generate responses only at F1 and F2.
However, the rectification that accompanies early visual processing
converts this contrast reversal into a more complex frequency-domain
representation with power predominantly at the second harmonics 2F1 and
2F2}

@@

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the LGN (Levitt
\emph{et al.}, 2001) and cortex (Hubel \& Wiesel, 1959; Movshon, 1975),
will introduce second harmonic components even when the stimulus itself
is modulated in an on-off fashion. Analogously, in the spatial domain,
so-called `simple cells' are sensitive to the polarity of a spatial
contrast modulation while `complex cells' respond to the presence of
patterned spatial contrast irrespective of its spatial position. The
SSVEP response to a contrast-reversing sine-wave grating therefore
contains information about nonlinear computations performed across a
range of retinal and cortical cell types.

@@

\includegraphics{review_files/figure-pdf/fig-contrastresponsefunction-output-1.pdf}

@@

\caption{\label{fig-contrastresponsefunction}Non-linear contrast
response function resulting from the population average of two full-wave
rectified signals passing through a non-linear transducer. Note the
`supersaturation' at high contrast levels.}

@@

One of the simplest nonlinearities is the function that describes a
cell's response to different levels of contrast. In
Figure~\ref{fig-contrastresponsefunction} this is modeled as described
above by a hyperbolic ratio function resulting in a saturating
non-linearity:

\begin{equation}
R_{\mathrm{max}} = \frac{C_{\mathrm{in}}^n}{C_{50}^n + C_{\mathrm{in}}^n},
\end{equation}

where \({R_{\mathrm{max}}}\) describes the maximum response level,
\(C_{\mathrm{in}}\) is the input contrast, \(C_{50}\) is the
`semi-saturation constant' (the point at which the response is at
half-maximum) and \emph{n} controls the steepness of the curve.

The hyperbolic ratio function is monotonic but the CRF resulting from
measuring the amplitude of the second harmonic (2F) component contains a
slight roll-off at high input contrasts. This results from the
distortion of the input sine waves at high contrast due to a combination
of the full-wave rectification and saturating non-linearity. Power at
other harmonics increases, and the total power generated by the input is
monotonic. This roll-off is often seen in experimental data and has been
referred to as `supersaturation' (Tyler \& Apkarian, 1985; Peirce,
2007).

To illustrate the effect of contrast gain control (Heeger, 1992), we
include a second component that contributes to the gain pool of the
first. At a single pair of (matched) contrast levels, we see a complex
pattern of intermodulation terms in the FT
(Figure~\ref{fig-intermodulation}).

\includegraphics{review_files/figure-pdf/fig-intermodulation-output-1.pdf}

\caption{\label{fig-intermodulation}Non-linear combination of two
independent contrast modulations at F1 and F2 results in
`intermodulation' terms that appear at low-order sums and differences of
those frequencies}

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex contains is the result of a
cascade of nonlinear retinal and cortical operations up to that point.
It therefore contains a `signature' of the computational nature, order
and parameters of those operations - including the shape of the
transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal.

This possibility was recognised in the early days of the technique
(Regan \& Regan, 1988). Although characterising the complete set of
computations along entire processing pathway is problematic, careful
parametric variation of the input stimuli does allow us to fit models of
early visual gain control using SSVEP data. \ldots{} MORE HERE -
Although this idea was suggested as early as 1988 (probably earlier) not
many people seem to have done it much. Candy \emph{et al.} (2001); Busse
\emph{et al.} (2009); Tsai \emph{et al.} (2012); Baker \& Wade (2017)
*** SEE ALSO RECENT WORK FROM WINAWER LAB - MUCH OF IT IS `FAST' VEPs
***

@@

\includegraphics{review_files/figure-pdf/fig-sweepvep-output-1.pdf}

@@

\caption{\label{fig-sweepvep}Sweep VEP simulation showing how a contrast
detection threshold can be estimated from sweep VEP data measured at
higher contrasts}

@@

Amblyopia is a disorder of binocular vision, characterised by one eye
contributing much less to perception than the other. This is often due
to strabismus (squint) or anisometropia (difference in optical
prescription between the eyes) during development. Contemporary accounts
suggest that the amblyopic eye is suppressed by signals from the fellow
eye. SSVEPs provide a convenient and objective method to characterise
the difference in neural response to signals in each eye, and typically
show reduced responses to stimuli in the amblyopic eye across the
contrast range (Baker \emph{et al.}, 2015; Lygo \emph{et al.}, 2021).
There are currently many novel binocular treatments for amblyopia under
development, often involving virtual reality or stereo display systems
designed to encourage the two eyes to work together. The steady-state
approach may be more sensitive and objective than typical acuity
measurements, and also has the potential to measure suppression between
the eyes directly (e.g. Zheng \emph{et al.}, 2019; Hu \emph{et al.},
2023; Du \emph{et al.}, 2023).

@@

Parkinson's

Alex - it would be good to add some CRFs from Afsari \emph{et al.}
(2014) to the figure. Is it possible to extract either the fitted
parameter values or the raw data for some control flies and some
mutants? Whatever you think is most compelling.

@@

\includegraphics{review_files/figure-pdf/fig-plotclinical-output-1.pdf}

@@

\caption{\label{fig-plotclinical}Example contrast response functions for
different clinical conditions. Panel (a) shows modelled contrast
response functions for epilepsy patients (blue) and control participants
(black), based on the data of Tsai et al.~(2011). Panel (b) shows
functions for the amblyopic (red) and fellow (black) eyes of adults with
amblyopia, based on the data of Baker et al.~(2015). Panel (c) shows
functions for children with (green) and without (black) a diagnosis of
autism, based on the data of Vilidaite et al.~(2018). Panel (d)}

@@

@@

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-Di-Russo2001}
Di Russo F, Spinelli D \& Morrone MC (2001).
\href{https://doi.org/10.1016/s0042-6989(01)00134-1}{Automatic gain
control contrast mechanisms are modulated by attention in humans:
Evidence from visual evoked potentials}. \emph{Vision Res} \textbf{41,}
2435--2447.

@@

>>>>

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for at least 100 years. One of the most
productive experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high SNR, and they allow us to analyze interactions between
stimulus components using a convenient mathematical framework. Here we
describe how SSVEPs have been used to study visual contrast over the
past 70 years (Dawson, 1954). Because our thinking about SSVEPs is
well-described by simple mathematical models, we embed code that
illustrates key steps in the modelling and analysis. This paper can
therefore be used both as a review of the use of SSVEP in measuring
human contrast processing, and as an interactive learning aid.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well (McKeefry \emph{et al.}, 1996;
Baseler \& Sutter, 1997; Di Russo \emph{et al.}, 2001\emph{a}).

Contrast is relatively simple to define: typically, it is specified as
the percentage deviation of a uniform stimulus from the background. So,
for example, a disk of 100 units of cone activation
(\(I_{\mathrm{stim}}\)) surrounded by a `background' of 50 units of
activation (\(I_{\mathrm{background}}\)) has a contrast of
\(\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}\)
= 100\%. Where patterns are more complex (for example, the sine-wave
gratings or Gabor patches common in vision science), the Michelson
(1927) definition of contrast is specified by the maximum and minimum
excursions from the mean:

\begin{equation}\phantomsection\label{eq-michelson}{\frac {I_{\mathrm{stimmax} }-I_{\mathrm{stimmin} }}{I_{\mathrm{stimmax} }+I_{\mathrm{stimmin} }}.}\end{equation}

@@

\includegraphics{Figures/centresurround.pdf}

@@

\caption{\label{fig-centresurround}The perceived contrast of a stimulus
depends on its context. A high contrast surround reduces the apparent
contrast of the central `probe' region.}

@@

Because it provides a direct read-out of neural population activity, the
SSVEP signal can reveal key features of neural signal transduction. For
example, by varying the peak stimulus contrast parametrically, a
`contrast vs response' function can be measured - where the `response'
is typically defined as the amplitude of the SSVEP frequency component
at the stimulus frequency, or a low multiple thereof. This corresponds
closely to similar functions reported by studies measuring single unit
activity or local field potentials in the cortex (Shapley \& Victor,
1980; Morrone \emph{et al.}, 1982). However the SSVEP has the advantage
that it is non-invasive, and so can be measured in awake, behaving human
participants.

@@

To better understand these processes, we will show how sinusoidal input
signals might be processed by the visual system to produce SSVEPs.
Figure~\ref{fig-linear} illustrates how sine waves of different
contrasts are processed in a linear system. The first panel shows the
input sine wave, which would be used to modulate stimulus amplitude over
time. Notice that there are five peaks in the waveform during the one
second sample, so the stimulation frequency is 5Hz (F1). The second
panel shows Fourier transform of the waveform, which contains a
substantial peak at this frequency. If we parametrically vary the
stimulus contrast (i.e.~the amplitude of the waveform), the amplitude of
the F1 component increases linearly with contrast (right panel).

@@

\includegraphics{Figures/linearsystem.pdf}

@@

\caption{\label{fig-linear}Illustration of a sinusoidal input signal
(left), its Fourier spectrum (middle), and how the amplitude of the
first harmonic component increases with contrast (right). These
calculations assume an entirely linear system.}

@@

Next we can consider the impact of a nonlinearity in processing on the
responses. One of the simplest nonlinearities is the function that
describes a cell's response to different levels of contrast. In
Figure~\ref{fig-transducer} this is modeled by a hyperbolic ratio
function resulting in a saturating non-linearity:

\begin{equation}\phantomsection\label{eq-nakarushton}{R_{\mathrm{max}} = \frac{C_{\mathrm{in}}^n}{C_{50}^n + C_{\mathrm{in}}^n},}\end{equation}

where \({R_{\mathrm{max}}}\) describes the maximum response level,
\(C_{\mathrm{in}}\) is the input contrast (or the time-varying
waveform), \(C_{50}\) is the `semi-saturation constant' (the point at
which the response is at half-maximum) and \emph{n} controls the
steepness of the curve (with a typical value around \(n=2\).

This nonlinearity has clear effects on signal transduction. The
sinusoidal waveform is distorted by the nonlinearity, and appears
frequency doubled (left panel of Figure~\ref{fig-transducer}; note that
other nonlinearities such as rectification and squaring can have similar
effects). The distortion is reflected in the Fourier spectrum (middle
panel of Figure~\ref{fig-transducer}), which now includes responses at
integer multiples (known as harmonics) of the stimulation frequency (10,
20 and 30Hz), but no response at the fundamental (5Hz). Finally, the
contrast response function is now nonlinear (right panel of
Figure~\ref{fig-transducer}). Although the hyperbolic ratio function
(Equation~\ref{eq-nakarushton}) is monotonic, the CRF resulting from
measuring the amplitude of the second harmonic (2F) component contains a
slight roll-off at high input contrasts. This results from the
distortion of the input sine waves at high contrast due to a combination
of the full-wave rectification and saturating non-linearity. Power at
other harmonics increases, and the total power generated by the input is
monotonic. This roll-off is often seen in experimental data and has been
referred to as `supersaturation' (Tyler \& Apkarian, 1985; Peirce,
2007).

@@

\includegraphics{Figures/transducer.pdf}

@@

\caption{\label{fig-transducer}Illustration of how a sinusoidal input
signal is affected by a nonlinear transducer. The input wavefore (left
panel) is frequency doubled by the squaring operator, and now features
energy at the second harmonic (10Hz) and its multiple (middle panel).
The contrast response function at the second harmonic is now nonlinear,
and saturates at high contrasts (right panel).}

@@

To illustrate the effect of contrast gain control (Heeger, 1992), we
next include a second component (a `mask') that contributes to the gain
pool of the first (the `target'). The mask stimulus will suppress the
response at the target frequency, reducing its amplitude. The
suppression is reciprocal - activity at the mask frequency is also
reduced by the presence of the target, in a contrast-dependent manner
(Busse \emph{et al.}, 2009). At a single pair of (matched) contrast
levels, we see a complex pattern of intermodulation terms in the
transduced waveforms (left panel of Figure~\ref{fig-twoinputs}) and in
the Fourier spectrum (middle panel of Figure~\ref{fig-twoinputs}). The
two components interact generating nonlinear intermodulation terms at
sums and differences of the input frequencies (F1, F2). The nature of
the interaction --- and therefore the pattern of intermodulation terms
--- is determined by the computations happening as the contrast signal
moves from retina to cortex (Regan \& Regan, 1988).

@@

\includegraphics{Figures/twoinputs.pdf}

@@

\caption{\label{fig-twoinputs}Demonstration of how two inputs of
different frequencies are processed by a nonlinear transducer.
Additional harmonics are apparent in the Fourier spectrum (middle
panel), and the contrast response function is suppressed (right panel).}

@@

If the two inputs were simply added together, the representation of the
resulting signal in the Fourier domain would be the linear sum of the
two independent signals (i.e.~peaks at F1 and F2). However, it is
important to take physiology into account. Contrast gain control reduces
the amplitude of the responses via a suppressive process, which can be
incorporated into our transducer function via an additional denominator
term:

\begin{equation}\phantomsection\label{eq-gaincontrol}{R_{\mathrm{max}} = \frac{C_{\mathrm{in}}^n}{C_{50}^n + C_{\mathrm{in}}^n + C_{\mathrm{mask}}^n},}\end{equation}

where \(C_\mathrm{mask}\) reflects the contrast of the mask component at
a distinct frequency from that of the target. The effect of this extra
term is to reduce the target response (see right panel of
Figure~\ref{fig-twoinputs}). For a linear contrast axis, the contrast
response function becomes shallower, whereas on a logarithmic contrast
axis it maintains its steepness and shifts to the right. Suppressive
effects of this kind have been obtained using SSVEP with a variety of
different types of mask, including orthogonal overlaid masks, surround
masks, and dichoptic masks (Burr \& Morrone, 1987; Ross \& Speed, 1991;
Candy \emph{et al.}, 2001; Busse \emph{et al.}, 2009; Cunningham
\emph{et al.}, 2017; Salelkar \& Ray, 2020; for a meta-analysis see
Baker \emph{et al.}, 2021).

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the LGN (Levitt
\emph{et al.}, 2001) and cortex (Hubel \& Wiesel, 1959; Movshon, 1975),
will introduce second harmonic components even when the stimulus itself
is modulated in an on-off fashion. Analogously, in the spatial domain,
so-called `simple cells' are sensitive to the polarity of a spatial
contrast modulation while `complex cells' respond to the presence of
patterned spatial contrast irrespective of its spatial position (Hubel
\& Wiesel, 1962). The SSVEP response to a contrast-reversing sine-wave
grating therefore contains information about nonlinear computations
performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' of the computational nature, order and
parameters of those operations - including the shape of the transducer
functions and the computations involved in signal combination. In
principle, that information could be recovered from the SSVEP signal - a
possibility recognised in the early days of the technique (Regan \&
Regan, 1988). Although characterising the complete set of computations
along entire processing pathway is challenging, careful parametric
variation of the input stimuli does allow us to fit models of early
visual gain control using SSVEP data. For example, Tsai \emph{et al.}
(2012) demonstrated that a gain control model gave a good account of the
pattern of intermodulation responses produced by two overlaid patterns
flickering at different frequencies. This was achieved by passing full
stimulus waveforms through the transducer nonlinearity, and calculating
the Fourier spectrum of the model output. Our own work on signal
combination across eyes and space similarly demonstrated close
correspondence between the predictions of a computational model and
empirical data in humans (Baker \& Wade, 2017). More detailed modelling
of intracortical recordings (Groen \emph{et al.}, 2022) has revealed
details of the timecourse of gain control effects, specifically that
normalization is delayed slightly relative to the initial visual
response.

Similar changes to the contrast response function might also be obtained
using adaptation paradigms, in which the visual system is exposed to
high contrast stimuli for long durations. Psychophysically, adaptation
increases detection thresholds, but has little effect on contrast
discrimination performance (Ross \emph{et al.}, 1993), much like pattern
masks (Foley, 1994). Although SSVEP adaptation effects show strong
tuning for orientation (Campbell \& Maffei, 1970; Vergeer \emph{et al.},
2018) and spatial frequency (Mecacci \& Spinelli, 1976), there appear to
be no studies measuring the full SSVEP contrast response function before
and after an extended period of adaptation.

@@

\includegraphics{Figures/sweepplot.pdf}

@@

\caption{\label{fig-sweepvep}Sweep VEP simulation showing how a contrast
detection threshold can be estimated from sweep VEP data measured at
higher contrasts. The solid line is the regression fit to the lowest
four data points, and the dashed line extrapolates the fit back to
determine the contrast value when y=0, which gives a threshold estimate
(red point).}

@@

Amblyopia is a disorder of binocular vision, characterised by one eye
contributing much less to perception than the other. This is often due
to strabismus (squint) or anisometropia (difference in optical
prescription between the eyes) during development. Contemporary accounts
suggest that the amblyopic eye is suppressed by signals from the fellow
eye. SSVEPs provide a convenient and objective method to characterise
the difference in neural response to signals in each eye, and typically
show reduced responses to stimuli in the amblyopic eye (see
Figure~\ref{fig-plotclinical}b) across the contrast range (Baker
\emph{et al.}, 2015; Lygo \emph{et al.}, 2021). There are currently many
novel binocular treatments for amblyopia under development, often
involving virtual reality or stereo display systems designed to
encourage the two eyes to work together. The steady-state approach may
be more sensitive and objective than typical acuity measurements, and
also has the potential to measure suppression between the eyes directly
(e.g. Zheng \emph{et al.}, 2019; Hu \emph{et al.}, 2023; Du \emph{et
al.}, 2023).

@@

Recent work on understanding Parkinson's disease has also used
\emph{Drosophila} genetic models. Afsari \emph{et al.} (2014) found that
mutant flies produced stronger SSVEP responses to flickering lights than
control flies (see Figure~\ref{fig-plotclinical}d). The authors
theorised that differences in early gain control during development
might lead to visual deficits later in life. Although visual responses
are a convenient assay of neural function, it is likely that same
general process applies throughout the whole brain, including in the
motor system where the core Parkinson's symptoms (tremor, rigidity, slow
movement) manifest. The SSVEP differences were reduced by a kinase
inhibitor that targets the dopamine system, demonstrating how model
organisms can be used to test new pharmacological treatments. SSVEP
responses also provide a potential method to diagnose Parkinson's before
any symptoms manifest, and to monitor the effect of treatments.

@@

\includegraphics{Figures/clinicalplot.pdf}

@@

\caption{\label{fig-plotclinical}Example contrast response functions for
different clinical conditions. Panel (a) shows modelled contrast
response functions for epilepsy patients (blue) and control participants
(black), based on the data of Tsai et al.~(2011). Panel (b) shows
functions for the amblyopic (red) and fellow (black) eyes of adults with
amblyopia, based on the data of Baker et al.~(2015). Panel (c) shows
functions for children with (green) and without (black) a diagnosis of
autism, based on the data of Vilidaite et al.~(2018). Panel (d) shows
data from \emph{Drosophila melanogaster} (fruit flies) from Afsari et
al.~(2014). One day-old flies expressing a human gene linked to
Parkinson's (hLRRK-G2019S) show increased SSVEP response amplitude and
sensitivity (red) compared to control animals (black).}

@@

\section{Attention}\label{attention}

Attention exerts a profound influence on visual performance: for
example, allowing people to attend covertly to a spatial location for
example, improves their performance on a target detection task
significantly (Bashinski \& Bacharach, 1980; Posner, 1980; Carrasco
\emph{et al.}, 2000; Cameron \emph{et al.}, 2002; Morrone \emph{et al.},
2004; Pestilli \emph{et al.}, 2009). In principle, this enhancement may
be driven both by modulation of the underlying signal or noise
characteristics, or by additional decision-theoretic factors such as
reduction in spatial uncertainty (Petrov \emph{et al.}, 2006; Gould
\emph{et al.}, 2007). Early experiments with non-human primates showed
little evidence for attentionally-driven changes in neuronal spike rates
(Luck \emph{et al.}, 1997; McAdams \& Maunsell, 1999; Mehta \emph{et
al.}, 2000\emph{a}, 2000\emph{b}; Marcus \& Van Essen, 2002), but with
the advent of spatially-resolved human brain imaging methods in the late
1990s it became apparent that spatial attention was linked to frank
changes in both fMRI BOLD responses (Tootell \emph{et al.}, 1998; Somers
\emph{et al.}, 1999; Gandhi \emph{et al.}, 1999; Brefczynski \& DeYoe,
1999; Kastner \emph{et al.}, 1999; Buracas \& Boynton, 2007; Silver
\emph{et al.}, 2007; Li \emph{et al.}, 2008; Murray, 2008) and EEG
signals (Ding \emph{et al.}, 2006).

Electrophysiological and psychophysical measurements of attentional
effects on both luminance and chromatic contrast strongly implicated
gain control an an underlying mechanism (Lu \& Dosher, 1998; Di Russo \&
Spinelli, 1999\emph{a}; Di Russo \emph{et al.}, 2001\emph{b}) and
demonstrated that these effects differed between chromatic and
achromatic pathways (Di Russo \& Spinelli, 1999\emph{b}). In the late
2000s a comprehensive theoretical model for attentional modulation was
developed that framed it as a gain control computation (Reynolds \&
Heeger, 2009; Boynton, 2009). This framework has proven to be
influential - explaining a wide range of phenomena from the earlier
literature and demonstrating subtle interactions between the size of the
attentional `spotlight' and the stimulus configuration which rationalise
many apparent contradictions in the literature. Subsequent papers
(effect of stimulus configuration critical). Then\ldots{} talk about
Lauritzen \emph{et al.} (2010) Verghese \emph{et al.} (2012) (1 other
Verghese paper) maybe mention Highsmith \& Crognale (2010), Wang \& Wade
(2011) and Martinovic \& Andersen (2018) Probably lots more Carasco
\ldots{}

@@

One widespread recent application of the SSVEP technique is in the
design of brain-computer interfaces, which seek to control some aspect
of a computer using neural signals. The high SNR and frequency
resolution of SSVEPs make them an ideal candidate for this approach.
Typical studies may involve presenting an array of stimuli at different
flicker frequencies, and having the participant select one either by
overt attention (i.e.~shifting fixation to foveate the selected
stimulus) or covert attention (i.e.~deploying attention to one stimulus
whilst keeping fixated). EEG signals are highly sensitive to both visual
field position (Ales \emph{et al.}, 2010) and attentional state
(Lauritzen \emph{et al.}, 2010; Verghese \emph{et al.}, 2012), so the
response to the selected stimulus will typically increase relative to
the others, allowing it to be identified by an on-line algorithm.

Probably a summary of BCIs here.

@@

\section{References}\label{references}

@@

\bibitem[\citeproctext]{ref-Ales2010}
Ales JM, Yates JL \& Norcia AM (2010).
\href{https://doi.org/10.1016/j.neuroimage.2010.05.016}{V1 is not
uniquely identified by polarity reversals of responses to upper and
lower visual field stimuli}. \emph{Neuroimage} \textbf{52,} 1401--1409.

@@

\bibitem[\citeproctext]{ref-Baker2021}
Baker DH, Vilidaite G \& Wade AR (2021).
\href{https://doi.org/10.1371/journal.pcbi.1009507}{Steady-state
measures of visual suppression}. \emph{PLoS computational biology}
\textbf{17,} e1009507.

@@

\bibitem[\citeproctext]{ref-Bashinski1980}
Bashinski HS \& Bacharach VR (1980). Enhancement of perceptual
sensitivity as the result of selectively attending to spatial locations.
\emph{Perception \& psychophysics} \textbf{28,} 241--248.

@@

\bibitem[\citeproctext]{ref-Boynton2009}
Boynton GM (2009). \href{https://doi.org/10.1016/j.visres.2008.11.001}{A
framework for describing the effects of attention on visual responses}.
\emph{Vision Res} \textbf{49,} 1129--1143.

@@

\bibitem[\citeproctext]{ref-Brefczynski1999}
Brefczynski JA \& DeYoe EA (1999). \href{https://doi.org/10.1038/7280}{A
physiological correlate of the 'spotlight' of visual attention}.
\emph{Nat Neurosci} \textbf{2,} 370--374.

\bibitem[\citeproctext]{ref-Buracas2007}
Buracas GT \& Boynton GM (2007).
\href{https://doi.org/10.1523/JNEUROSCI.3162-06.2007}{The effect of
spatial attention on contrast response functions in human visual
cortex}. \emph{J Neurosci} \textbf{27,} 93--97.

\bibitem[\citeproctext]{ref-Burr1987}
Burr DC \& Morrone MC (1987).
\href{https://doi.org/10.1113/jphysiol.1987.sp016643}{Inhibitory
interactions in the human vision system revealed in pattern-evoked
potentials}. \emph{J Physiol} \textbf{389,} 1--21.

@@

\bibitem[\citeproctext]{ref-Cameron2002}
Cameron EL, Tai JC \& Carrasco M (2002).
\href{https://doi.org/10.1016/s0042-6989(02)00039-1}{Covert attention
affects the psychometric function of contrast sensitivity}. \emph{Vision
Res} \textbf{42,} 949--967.

@@

\bibitem[\citeproctext]{ref-Campbell1970}
Campbell FW \& Maffei L (1970).
\href{https://doi.org/10.1113/jphysiol.1970.sp009085}{Electrophysiological
evidence for the existence of orientation and size detectors in the
human visual system}. \emph{J Physiol} \textbf{207,} 635--652.

@@

\bibitem[\citeproctext]{ref-Carrasco2000}
Carrasco M, Penpeci-Talgar C \& Eckstein M (2000).
\href{https://doi.org/10.1016/s0042-6989(00)00024-9}{Spatial covert
attention increases contrast sensitivity across the CSF: Support for
signal enhancement}. \emph{Vision Res} \textbf{40,} 1203--1215.

@@

\bibitem[\citeproctext]{ref-Cunningham2017}
Cunningham DGM, Baker DH \& Peirce JW (2017).
\href{https://doi.org/10.1167/17.5.10}{{Measuring nonlinear signal
combination using EEG}}. \emph{Journal of Vision} \textbf{17,} 10--10.

@@

\bibitem[\citeproctext]{ref-DiRusso1999}
Di Russo F \& Spinelli D (1999\emph{a}).
\href{https://doi.org/10.1016/S0042-6989(99)00031-0}{Electrophysiological
evidence for an early attentional mechanism in visual processing in
humans}. \emph{Vision Research} \textbf{39,} 2975--2985.

\bibitem[\citeproctext]{ref-DiRusso1999b}
Di Russo F \& Spinelli D (1999\emph{b}). Spatial attention has different
effects on the magno-and parvocellular pathways. \emph{Neuroreport}
\textbf{10,} 2755--2762.

\bibitem[\citeproctext]{ref-DiRusso2001}
Di Russo F, Spinelli D \& Morrone MC (2001\emph{b}).
\href{https://doi.org/10.1016/S0042-6989(01)00134-1}{Automatic gain
control contrast mechanisms are modulated by attention in humans:
Evidence from visual evoked potentials}. \emph{Vision Research}
\textbf{41,} 2435--2447.

\bibitem[\citeproctext]{ref-Di-Russo2001}
Di Russo F, Spinelli D \& Morrone MC (2001\emph{a}).
\href{https://doi.org/10.1016/s0042-6989(01)00134-1}{Automatic gain
control contrast mechanisms are modulated by attention in humans:
Evidence from visual evoked potentials}. \emph{Vision Res} \textbf{41,}
2435--2447.

\bibitem[\citeproctext]{ref-Ding2006}
Ding J, Sperling G \& Srinivasan R (2006). Attentional modulation of
{SSVEP} power depends on the network tagged by the flicker frequency.
\emph{Cerebral cortex} \textbf{16,} 1016--1029.

@@

\bibitem[\citeproctext]{ref-Gandhi1999}
Gandhi SP, Heeger DJ \& Boynton GM (1999).
\href{https://doi.org/10.1073/pnas.96.6.3314}{Spatial attention affects
brain activity in human primary visual cortex}. \emph{Proc Natl Acad Sci
U S A} \textbf{96,} 3314--3319.

\bibitem[\citeproctext]{ref-Gould2007}
Gould IC, Wolfgang BJ \& Smith PL (2007).
\href{https://doi.org/10.1167/7.13.4}{Spatial uncertainty explains
exogenous and endogenous attentional cuing effects in visual signal
detection}. \emph{J Vis} \textbf{7,} 4.1--17.
